---
title: "Predicting Billboard #1 Hits"
output: pdf_document
date: "Authors: May Eindra Tet Toe, Anastasia Tountas, Tran Anh Thu Phung, Harry Nguyen"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary
Insert a summary of your analysis here.

## Introduction
Insert an introduction to your analysis here.

## Methods and Results

### Overview of our Methodology
The Billboard Hot 100 dataset provides us with two datasets: `billboard.csv`, 
which contains one row per song that reached #1 along with it's features, and `topics.csv`,
a reference list of lyrical theme labels (`lyrical_topics`). Our team decided to 
combine the two datasets to `billboard_clean` in order to create more comprehensive data 
to better equip our model. Here is an overview of our data analysis methodology:

1. **Reading & Wrangling:** read and wrangle the dataset into one tidy combined dataset

2. **Exploratory Data Analysis:** explore the data to understand the distribution of features and their relationships with the target variable

3. **Model Building:** build a regression model to predict the number of weeks a song stays at #1 based on its features

4. **Model Evaluation:** evaluate the performance of our model using appropriate metrics and validation techniques

5. **Results and Conclusion:** summarize our findings, discuss the implications of our results, and suggest potential future work or improvements to our analysis

### Load libraries and set seed
```{r load-libraries}
library(tidyverse)
library(tidytuesdayR)
library(knitr)
library(dplyr)
```
```{r set-seed}
set.seed(123)
```
### Reading & Wrangling our dataset from the web into R

- The two datasets were joined using a `left_join()` on `lyrical_topic` (billboard) 
  and `lyrical_topics`(topics).
- **Identifier columns** (`song`, `artist`, `date`) were removed as they are
  not predictive features.
- **`non_consecutive`** was removed to prevent data leakage — it can only
  be determined after a song has completed its full chart run.
- **`rating_1`, `rating_2`, `rating_3`** were removed in favour of
  `overall_rating` (their mean) to avoid multicollinearity.
- **Free-text columns** were removed avoid the high dimensionality 
  and noise associated with Natural Language Processing:
 `lyrics`, `u_s_artwork`, `sound_effects`, `song_structure`,
  `featured_artists`, `songwriters`,
  `songwriters_w_o_interpolation_sample_credits`, `producers`,
  `double_a_side`, and `talent_contestant`.
- **High-cardinality columns** removed because they generalise poorly:
  `label`, `parent_label`, and `artist_place_of_origin`.
- **Redundant columns** were removed: `keys` is replaced by `simplified_key`;
  `cdr_style` and `discogs_style` are replaced by `cdr_genre` and `discogs_genre` respectively.
- **`discogs_genre`** was removed in favour of `cdr_genre` because cdr has more reliable 
  class definitions, thus improving internal validity.
- **Ambiguous columns** were removed: `featured_in_a_then_contemporary_play`,
  `featured_in_a_then_contemporary_film`, and
  `featured_in_a_then_contemporary_t_v_show`.
- All remaining character columns were converted to factors using
  `as.factor()`, and rows with any missing values were dropped with
  `drop_na()`.

```{r load-data}
# Load the dataset
tuesdata <- tidytuesdayR::tt_load(2025, week = 34)
# Extract the datasets
billboard <- tuesdata$billboard
topics <- tuesdata$topics
```
```{r join-data}
# Join the two datasets
billboard_joined <- billboard |>
  left_join(topics, by = c("lyrical_topic" = "lyrical_topics"))
```
```{r wrangle-data}
billboard_clean <- billboard_joined |>
  # Remove identifier columns
  select(-c(song, artist, date)) |>
  # Remove data-leakage column
  select(-non_consecutive) |>
  # Remove individual judge scores to avoid multicollinearity
  select(-c(rating_1, rating_2, rating_3)) |>
  # Remove free-text columns
  select(-c(
    lyrics,
    u_s_artwork,
    sound_effects,
    song_structure,
    featured_artists,
    songwriters,
    songwriters_w_o_interpolation_sample_credits,
    producers,
    double_a_side,
    talent_contestant
  )) |>
  # Remove high-cardinality columns
  select(-c(label, parent_label, artist_place_of_origin)) |>
  # Remove redundant columns and discogs_genre
  select(-c(cdr_style, discogs_style, keys, discogs_genre)) |>
  # Remove ambiguous columns
  select(-c(
    featured_in_a_then_contemporary_play,
    featured_in_a_then_contemporary_film,
    featured_in_a_then_contemporary_t_v_show
  )) |>
  # Convert remaining character columns to factors
  mutate(across(where(is.character), as.factor)) |>
  drop_na()
```

### Exploratory Data Analysis
We first examine the distribution of our target variable,
`weeks_at_number_one`, to check for skewness.

```{r eda-target-dist, fig.cap="Distribution of weeks at number one."}
ggplot(billboard_clean, aes(x = weeks_at_number_one)) +
  geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
  labs(
    title = "Distribution of Weeks at Number One",
    x = "Weeks at Number One",
    y = "Count"
  ) +
  theme_minimal()
```

The distribution is heavily right-skewed. The majority of songs spent only
1–3 weeks at #1, with counts dropping off sharply afterwards. A small
number of songs stayed for 10 or more weeks, forming a long right tail.

Next, we examine the distributions of the five Spotify audio features, to
identify any skewness that may require transformation before modelling.

```{r eda-numeric-distributions, fig.cap="Distributions of the five Spotify audio features."}
billboard_clean |>
  select(bpm, energy, danceability, happiness, acousticness) |>
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(fill = "steelblue", colour = "white", bins = 30) +
  facet_wrap(~feature, scales = "free") +
  labs(
    title = "Distributions of Spotify Audio Features",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal()
```

`acousticness` is heavily right-skewed, with most songs clustered near
zero. `happiness` shows a bimodal pattern with peaks around 30 and around 80.
`energy` is very slightly left-skewed, while `bpm` and `danceability` 
are roughly normal.

Finally, we compare average `weeks_at_number_one` by genre to see whether
certain genres are associated with longer chart runs.

```{r eda-genre, fig.cap="Average weeks at number one by genre."}
billboard_clean |>
  group_by(cdr_genre) |>
  summarise(mean_weeks = mean(weeks_at_number_one), .groups = "drop") |>
  ggplot(aes(
    x = reorder(cdr_genre, mean_weeks),
    y = mean_weeks
  )) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Average Weeks at Number One by Genre",
    x = "Genre",
    y = "Average Weeks at Number One"
  ) +
  theme_minimal()
```

Electronic/Dance;Latin seems to have the highest average (~10 weeks).
Pop;Reggae and Folk/Country;March also rank highly. Mainstream genres 
like Pop and Rock sit in the middle of the range. At the lower end, we
can find niche genres like Electronic/Dance;Funk/Soul, averaging to 1 week.

### Feature Transformations

Based on the Exploratory Data Analysis, three variables require transformation before modelling:

- **`weeks_at_number_one`** is heavily right-skewed
- **`acousticness`** is also heavily right-skewed
- **`happiness`** is mildly left-skewed

Therefore, we apply log transformation to these three variables to compresses the tail and produce
a more symmetric target distribution. We will be using `log1p()` instead of `log()` to handle any 
zero values in the data.
