---
title: "Predicting Billboard #1 Hits"
output: pdf_document
date: "Authors: May Eindra Tet Toe, Anastasia Tountas, Tran Anh Thu Phung, Harry Nguyen"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Summary
Insert a summary of your analysis here.

### Introduction
Insert an introduction to your analysis here.

### Methods and Results

#### Overview of our Methodology
The Billboard Hot 100 dataset provides us with two datasets: `billboard.csv`, 
which contains one row per song that reached #1 along with it's features, and `topics.csv`,
a reference list of lyrical theme labels (`lyrical_topics`). Our team decided to 
combine the two datasets to `billboard_clean` in order to create more comprehensive data 
to better equip our model. Here is an overview of our data analysis methodology:

1. **Reading & Wrangling:** read and wrangle the dataset into one tidy combined dataset

2. **Exploratory Data Analysis:** explore the data to understand the distribution of features and their relationships with the target variable

3. **Model Building:** build a regression model to predict the number of weeks a song stays at #1 based on its features

4. **Model Evaluation:** evaluate the performance of our model using appropriate metrics and validation techniques

5. **Results and Conclusion:** summarize our findings, discuss the implications of our results, and suggest potential future work or improvements to our analysis

#### Load libraries and set seed
```{r load-libraries}
library(tidyverse)
library(tidytuesdayR)
library(knitr)
library(dplyr)
```
```{r set-seed}
set.seed(123)
```
#### Reading & Wrangling our dataset from the web into R

- The two datasets were joined using a `left_join()` on `lyrical_topic` (billboard) 
  and `lyrical_topics`(topics).
- **Identifier columns** (`song`, `artist`, `date`) were removed as they are
  not predictive features.
- **`non_consecutive`** was removed to prevent data leakage — it can only
  be determined after a song has completed its full chart run.
- **`rating_1`, `rating_2`, `rating_3`** were removed in favour of
  `overall_rating` (their mean) to avoid multicollinearity.
- **Free-text columns** were removed avoid the high dimensionality 
  and noise associated with Natural Language Processing:
 `lyrics`, `u_s_artwork`, `sound_effects`, `song_structure`,
  `featured_artists`, `songwriters`,
  `songwriters_w_o_interpolation_sample_credits`, `producers`,
  `double_a_side`, and `talent_contestant`.
- **High-cardinality columns** removed because they generalise poorly:
  `label`, `parent_label`, and `artist_place_of_origin`.
- **Redundant columns** were removed: `keys` is replaced by `simplified_key`;
  `cdr_style` and `discogs_style` are replaced by `cdr_genre` and `discogs_genre` respectively.
- **`discogs_genre`** was removed in favour of `cdr_genre` because cdr has more reliable 
  class definitions, thus improving internal validity.
- **Ambiguous columns** were removed: `featured_in_a_then_contemporary_play`,
  `featured_in_a_then_contemporary_film`, and
  `featured_in_a_then_contemporary_t_v_show`.
- All remaining character columns were converted to factors using
  `as.factor()`, and rows with any missing values were dropped with
  `drop_na()`.

```{r load-data}
# Load the TidyTuesday data for the week of August 26, 2025
tuesdata <- tidytuesdayR::tt_load(2025, week = 34)
# Extract the datasets
billboard <- tuesdata$billboard
topics <- tuesdata$topics
```
```{r join-data}
# Join billboard with the topics reference list on lyrical topic category
billboard_joined <- billboard |>
  left_join(topics, by = c("lyrical_topic" = "lyrical_topics"))
```
```{r wrangle-data}
billboard_clean <- billboard_joined |>
  # Remove identifier columns
  select(-c(song, artist, date)) |>
  # Remove data-leakage column
  select(-non_consecutive) |>
  # Remove individual judge scores to avoid multicollinearity
  select(-c(rating_1, rating_2, rating_3)) |>
  # Remove free-text columns
  select(-c(
    lyrics,
    u_s_artwork,
    sound_effects,
    song_structure,
    featured_artists,
    songwriters,
    songwriters_w_o_interpolation_sample_credits,
    producers,
    double_a_side,
    talent_contestant
  )) |>
  # Remove high-cardinality columns
  select(-c(label, parent_label, artist_place_of_origin)) |>
  # Remove redundant columns and discogs_genre
  select(-c(cdr_style, discogs_style, keys, discogs_genre)) |>
  # Remove ambiguous columns
  select(-c(
    featured_in_a_then_contemporary_play,
    featured_in_a_then_contemporary_film,
    featured_in_a_then_contemporary_t_v_show
  )) |>
  # Convert remaining character columns to factors
  mutate(across(where(is.character), as.factor)) |>
  drop_na()

glimpse(billboard_clean)
```
