---
title: "Predicting Billboard #1 Hits"
output: pdf_document
date: "Authors: May Eindra Tet Toe, Anastasia Tountas, Tran Anh Thu Phung, Harry Nguyen"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

Insert a summary of your analysis here.

## Introduction

Insert an introduction to your analysis here.

## Methods and Results

### Overview of our Methodology

The Billboard Hot 100 dataset provides us with two datasets: `billboard.csv`, which contains one row per song that reached #1 along with it's features, and `topics.csv`, a reference list of lyrical theme labels (`lyrical_topics`). Our team decided to combine the two datasets to `billboard_clean` in order to create more comprehensive data to better equip our model. Here is an overview of our data analysis methodology:

1.  **Reading & Wrangling:** read and wrangle the dataset into one tidy combined dataset

2.  **Exploratory Data Analysis:** explore the data to understand the distribution of features and their relationships with the target variable

3.  **Model Building:** build a regression model to predict the number of weeks a song stays at #1 based on its features

4.  **Model Evaluation:** evaluate the performance of our model using appropriate metrics and validation techniques

5.  **Results and Conclusion:** summarize our findings, discuss the implications of our results, and suggest potential future work or improvements to our analysis

### Load libraries and set seed

```{r load-libraries}
library(tidyverse)
library(tidytuesdayR)
library(knitr)
library(dplyr)
library(caret)
library(tidymodels)
```

```{r set-seed}
set.seed(123)
```

### Reading & Wrangling our dataset from the web into R

-   The two datasets were joined using a `left_join()` on `lyrical_topic` (billboard) and `lyrical_topics`(topics).
-   **Identifier columns** (`song`, `artist`, `date`) were removed as they are not predictive features.
-   **`non_consecutive`** was removed to prevent data leakage — it can only be determined after a song has completed its full chart run.
-   **`rating_1`, `rating_2`, `rating_3`** were removed in favour of `overall_rating` (their mean) to avoid multicollinearity.
-   **Free-text columns** were removed avoid the high dimensionality and noise associated with Natural Language Processing: `lyrics`, `u_s_artwork`, `sound_effects`, `song_structure`, `featured_artists`, `songwriters`, `songwriters_w_o_interpolation_sample_credits`, `producers`, `double_a_side`, and `talent_contestant`.
-   **High-cardinality columns** removed because they generalise poorly: `label`, `parent_label`, and `artist_place_of_origin`.
-   **Redundant columns** were removed: `keys` is replaced by `simplified_key`; `cdr_style` and `discogs_style` are replaced by `cdr_genre` and `discogs_genre` respectively.
-   **`discogs_genre`** was removed in favour of `cdr_genre` because cdr has more reliable class definitions, thus improving internal validity.
-   **Ambiguous columns** were removed: `featured_in_a_then_contemporary_play`, `featured_in_a_then_contemporary_film`, and `featured_in_a_then_contemporary_t_v_show`.
-   All remaining character columns were converted to factors using `as.factor()`, and rows with any missing values were dropped with `drop_na()`.

```{r load-data}
# Load the dataset
tuesdata <- tidytuesdayR::tt_load(2025, week = 34)
# Extract the datasets
billboard <- tuesdata$billboard
topics <- tuesdata$topics
```

```{r join-data}
# Join the two datasets
billboard_joined <- billboard |>
  left_join(topics, by = c("lyrical_topic" = "lyrical_topics"))
```

```{r wrangle-data}
billboard_clean <- billboard_joined |>
  # Remove identifier columns
  select(-c(song, artist, date)) |>
  # Remove data-leakage column
  select(-non_consecutive) |>
  # Remove individual judge scores to avoid multicollinearity
  select(-c(rating_1, rating_2, rating_3)) |>
  # Remove free-text columns
  select(-c(
    lyrics,
    u_s_artwork,
    sound_effects,
    song_structure,
    featured_artists,
    songwriters,
    songwriters_w_o_interpolation_sample_credits,
    producers,
    double_a_side,
    talent_contestant
  )) |>
  # Remove high-cardinality columns
  select(-c(label, parent_label, artist_place_of_origin)) |>
  # Remove redundant columns and discogs_genre
  select(-c(cdr_style, discogs_style, keys, discogs_genre)) |>
  # Remove ambiguous columns
  select(-c(
    featured_in_a_then_contemporary_play,
    featured_in_a_then_contemporary_film,
    featured_in_a_then_contemporary_t_v_show
  )) |>
  # Convert remaining character columns to factors
  mutate(across(where(is.character), as.factor)) |>
  drop_na()
```

### Exploratory Data Analysis

First, we find the summary statistics for our target variable, `weeks_at_number_one`, and our key numeric predictors: `overall_rating`, `bpm`, `energy`, `danceability`, `happiness`, `acousticness`, `loudness_d_b`, `front_person_age`, and `length_sec`.

```{r eda-summary, echo = FALSE}
billboard_clean |>
  select(
    weeks_at_number_one, overall_rating, bpm,
    energy, danceability, happiness,
    acousticness, loudness_d_b, front_person_age, length_sec
  ) |>
  pivot_longer(cols = everything(), names_to = "Variable") |>
  group_by(Variable) |>
  summarise(
    Mean   = round(mean(value), 2),
    SD     = round(sd(value), 2),
    Min    = round(min(value), 2),
    Median = round(median(value), 2),
    Max    = round(max(value), 2),
    .groups = "drop"
  ) |>
  kable(caption = "Summary statistics")
```

Then, we examine the distribution of our target variable, `weeks_at_number_one`, to check for skewness.

```{r eda-target-dist, fig.cap="Distribution of weeks at number one.", echo = FALSE}
ggplot(billboard_clean, aes(x = weeks_at_number_one)) +
  geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
  labs(
    title = "Distribution of Weeks at Number One",
    x = "Weeks at Number One",
    y = "Count"
  ) +
  theme_minimal()
```

The distribution is heavily right-skewed. The majority of songs spent only 1 to 3 weeks at #1, with counts dropping off sharply afterwards. A small number of songs stayed for 10 or more weeks, forming a long right tail.

Next, we examine the distributions of the five Spotify audio features, to identify any skewness that may require transformation before modelling.

```{r eda-numeric-distributions, fig.cap="Distributions of the five Spotify audio features.", echo = FALSE}
billboard_clean |>
  select(bpm, energy, danceability, happiness, acousticness) |>
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(fill = "steelblue", colour = "white", bins = 30) +
  facet_wrap(~feature, scales = "free") +
  labs(
    title = "Distributions of Spotify Audio Features",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal()
```

`acousticness` is heavily right-skewed, with most songs clustered near zero. `happiness` shows a bimodal pattern with peaks around 30 and around 80. `energy` is very slightly left-skewed, while `bpm` and `danceability` are roughly normal.

Finally, we compare average `weeks_at_number_one` by genre to see whether certain genres are associated with longer chart runs.

```{r eda-genre, fig.cap="Average weeks at number one by genre.", echo = FALSE}
billboard_clean |>
  group_by(cdr_genre) |>
  summarise(mean_weeks = mean(weeks_at_number_one), .groups = "drop") |>
  ggplot(aes(
    x = reorder(cdr_genre, mean_weeks),
    y = mean_weeks
  )) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Average Weeks at Number One by Genre",
    x = "Genre",
    y = "Average Weeks at Number One"
  ) +
  theme_minimal()
```

Electronic/Dance;Latin seems to have the highest average (\~10 weeks). Pop;Reggae and Folk/Country;March also rank highly. Mainstream genres like Pop and Rock sit in the middle of the range. At the lower end, we can find niche genres like Electronic/Dance;Funk/Soul, averaging to 1 week.

### Feature Transformations

Based on the Exploratory Data Analysis, two variables require transformation before modelling:

-   **`weeks_at_number_one`** is heavily right-skewed
-   **`acousticness`** is also heavily right-skewed

Therefore, we apply log transformation to these three variables to compresses the tail and produce a more symmetric target distribution. We will be using `log1p()` instead of `log()` to handle any zero values in the data.

```{r log-transformed}
billboard_log_transformed <- billboard_clean |> 
  mutate(
    log_weeks_at_number_one = log1p(weeks_at_number_one),
    log_acousticness = log1p(acousticness)
  )
```

Moreover, several songs in the dataset were associated with more than one musical genre (for example, `Electronic/Dance; Latin` or `Pop;Funk/Soul`). Treating these combined labels as single categorical values would incorrectly imply that each combination represents a distinct genre and would lead to sparse, poorly interpretable factor levels. To address this, the `cdr_gerne` variable was processed as a multi-label feature. Individual genre were first separated into their component parts and then encoded as binary indicator variables using one-hot encoding.

```{r one-hot-encoding for genre}
billboard_genres <- billboard_log_transformed |> 
  separate_rows(cdr_genre, sep = ";") |> 
  mutate (cdr_genre = str_trim(cdr_genre))

billboard_genres_wide <- billboard_genres |> 
  mutate(value = 1) |> 
  pivot_wider(
    names_from = cdr_genre, 
    values_from = value,
    values_fill = 0
  )
```

Transform the Simplified Key:
``` {r simplify-key}
billboard_simplify_key <- billboard_genres_wide |> 
  mutate(
    simplified_key = case_when(
      simplified_key == "Multiple Keys"          ~ "Multiple Keys",
      str_ends(simplified_key, "m")              ~ "Minor",  # Am, Bbm, Bm, Cm, etc.
      TRUE                                        ~ "Major"  # A, Ab, B, Bb, C, etc.
    ) |> as.factor()
  )

billboard_model_final <- billboard_simplify_key |> 
  select(-weeks_at_number_one, -acousticness)
```

### Feature Selection
#### Drop Redundant Columns
- Demographic identity columns: Race and gender of the artist or producing crews don't directly cause a song to stay longer at #1, so we should drop these columns.
- Songwriter/producer role overlaps
- Timing collinearity
- Lyric redundant content
- External content

```{r drop-manually}
columns_to_drop <- c(
  # Demographic identity
  "artist_male","artist_white", "artist_black", "songwriter_male", "songwriter_white", "producer_male", "producer_white", 
  # Artist structure
  "artist_structure", "group_named_after_non_lead_singer",
  # Songwriter/producer overlaps
  "artist_is_only_songwriter", "artist_is_only_producer", "songwriter_is_a_producer",
  # Timing
  "instrumental_length_sec",  "intro_length_sec",
  # Lyrical content 
  "lyrical_topic","lyrical_narrative",
  # External content
  "written_for_a_play", "written_for_a_film", "written_for_a_t_v_show", "eurovision_entry", "topped_the_charts_by_multiple_artist", "associated_with_dance")

billboard_manual_dropped <- billboard_model_final |>
  select(-all_of(columns_to_drop))
```

#### Drop columns with nearly zero variance :
```{r variance-test}
# Identify and remove near-zero variance predictors
nzv_cols <- nearZeroVar(billboard_manual_dropped, 
                        freqCut = 95/5,  # flag if most common value appears 19x more than second most common
                        uniqueCut = 5,   # flag if fewer than 5% unique values
                        names = TRUE)
nzv_cols

billboard_model_selected <- billboard_manual_dropped |>
  select(-all_of(nzv_cols))
```
### Regression Model 
``` {r train-test-split}
# Train/test split
billboard_split <- initial_split(billboard_model_selected, prop = 0.7)
train_data <- training(billboard_split)
test_data <- testing(billboard_split)
```

``` {r model-setup}
lm_recipe <- recipe(log_weeks_at_number_one ~ ., data = train_data) |>

  # Convert any remaining factor predictors to dummy variables
  # (simplified_key gets 3 dummies; reference level = "Major")
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>

  # Remove any predictor with near-zero variance that survived earlier filtering
  step_nzv(all_predictors()) |>

  # Centre and scale all numeric predictors for coefficient interpretability
  step_normalize(all_numeric_predictors()) |>

  # Drop any predictor that is a perfect linear combination of others
  step_lincomb(all_numeric_predictors())

# Inspect the processed training frame (sanity check)
prep(lm_recipe) |> bake(new_data = NULL) |> glimpse()

lm_spec <- linear_reg() |>
  set_engine("lm") |>
  set_mode("regression")

lm_workflow <- workflow() |>
  add_recipe(lm_recipe) |>
  add_model(lm_spec)

```

``` {r cross-validation}
set.seed(123)
cv_folds <- vfold_cv(
  train_data,
  v       = 5,
  repeats = 3,
  strata  = log_weeks_at_number_one
)

cv_results <- fit_resamples(
  lm_workflow,
  resamples = cv_folds,
  metrics   = metric_set(rmse, rsq, mae),
  control   = control_resamples(save_pred = TRUE)
)

# Cross-validation summary
cv_metrics <- collect_metrics(cv_results)
print(cv_metrics)
```

```{r model-fit}
final_fit <- fit(lm_workflow, data = train_data)
```

``` {r test_prediction}
test_predictions <- predict(final_fit, new_data = test_data) |>
  bind_cols(test_data |> select(log_weeks_at_number_one)) |>
  rename(
    predicted = .pred,
    actual    = log_weeks_at_number_one
  )

test_metrics <- test_predictions |>
  metrics(truth = actual, estimate = predicted)

print(test_metrics)
```
